### 2023-06-25 Draft.tex
Added additional pictures

This is where filtering is utilized to pass through and create a desired output image, catering to one' individual goals. The two major categories of filtering in Image processing are \emph{Linear} and \emph{Nonlinear filters}, with each
having variations of their own, tailored to their specific use cases, whether for noise suppresion, edge detection or image enhancement\cite{MISRA2020289}. The comparison between each filter is not only determining which filters produces the highest quality output, yet
also defined by its computing complexity, both in space and time complexity. Optimizing these filters by means of different algorithms and strategies while maintaining its quality is highly preferable.

In this article, we will discuss the differences between different types of linear and non linear filters.\footnotemark\footnotetext{Written by Jose}
\section{Linear Filters}
With \textbf{Linear Filtering}, or specifically linear spatial filtering, the function which is used to pass the image through
must be linear and shift invariant. The function $\pmb{L}$ is considered linear if there exists two constants $a$ and $b$ for any
two input pixels $f_1(m, n)$ and $f_2(m, n)$ such that

@@

The desired output pixel $g(i, j)$, where $i$ and $j$ specify the
coordinates of it, is based on a $M \times N$ sized neighborhood, 
meaning not only does one pixel define an output pixel, but also a specified number
of its neighbors. The influence of each pixel in the neighborhood is defined by the filter coefficient $h(k, l)$,
also called its \emph{kernel} or \emph{mask}\cite{Szeliski_2022}.

@@

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.30\textwidth]{codes/color.jpg}}
    \subfigure[shifted]{\includegraphics[width=0.30\textwidth]{img/colorShift.jpg}}
    \caption{Hue shifting using linear filters of the GALEX telescope image\cite{NASA_on_The_Commons_23}}
    \label{fig:hueshift}
\end{figure}\end{minipage}
\begin{itemize}
    \item \textbf{Box Filter:} each pixel is determined by the average of a $K \times K$ 
        neighborhood of pixels. It is also often called the \emph{moving average}, as the calculation can be done with scanlines, subtracting the
        old pixel and adding the newest pixels to the running sum. As the pixel is influenced equally by a square neighborhood, this may not be ideal as some `square artifacts' can be seen.
    \item \textbf{Bartlett Filter:} one solution to this is by separably convolving the image
        with a linear `tent' function (having its peak in the middle). A $3\times 3$ version of this is called the bilinear kernel.
    \item \textbf{Approximate Gaussian Filter:} if the Bartlett kernel is convolved with itself, this will produce an approximate version of the gaussian kernel.
\end{itemize}

@@

These filters are mainly used to smoothen(blur) the image, where the low frequencies are let through as well as reducing the effect of the high frequencies (high frequencies and low frequencies both referring to
the brightness/darkness of each pixel). In practical applications, this is done to reduce high frequency noises.
A creative application to smoothing the images, is that the smoothen images can be used as a mask to sharpen the image by adding the difference between the original and blurred image.
This method is called \emph{unsharp masking}\cite{Szeliski_2022}.
\[
  g_{sharp} = f + \gamma(f - h_{blur} \ast f)  
\]
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.19\textwidth]{img/original_mask.jpg}}
    \subfigure[unsharp mask]{\includegraphics[width=0.19\textwidth]{img/unsharp_mask.jpg}}
    \subfigure[difference]{\includegraphics[width=0.19\textwidth]{img/difference_unsharp.jpg}}
    \subfigure[sharpened]{\includegraphics[width=0.19\textwidth]{img/sharpened.jpg}}
    \caption{Unsharp masking with Gaussian blur}\label{fig:unsharp_mask}
\end{figure}\end{minipage}
The result of the image traces a bright outline througout the image between the darker(shadow) and lighter part. It can also be seen, that
the difference between the blurred image and the original results in an approximate edge detection. A better way to extract edges by means of linear filtering is
by the use of the \emph{Sobel filter}. \cite{Szeliski_2022}

@@

A similar to operator to the sobel filter is called the \emph{Prewitt operator}, which instead of convolving it with a perpendicular tent filter,
it instead uses an average operator.\cite{NIXON2020141}

>>>>

This is where filtering is utilized to pass through and create a desired output image, catering to one' individual goals. The two major categories of filtering in Image processing are \emph{Linear} and \emph{Nonlinear filters}, with each
having variations of their own, tailored to their specific use cases, whether for noise suppresion, edge detection or image enhancement\cite{MISRA2020289}. Not only are filters used in isolation, but they can aid
in further methods and algorithms in image processing, as they create aim to create clean and workable images for computers. The comparison between each filter is not only determining which filters produces the highest quality output, yet
also defined by its computing complexity, both in space and time complexity. Optimizing these filters by means of different algorithms and strategies while maintaining its quality is highly preferable.

The two major categories of filtering in image processing are \textbf{\emph{Linear Filters}} and \textbf{\emph{Non-linear Filters}}. Linear Filtering is a simpler, yet powerful strategy for image processing, as they
only require a linear neighborhood operation to calculate the output image. For difficult scenarios however, it may not be possible to only rely on basic multiplications and sums. This is where Non-linear filters comes in handy, as they
feature concepts such as cellular automata. This article aims to discuss different types of filters, their applications and how each of them provides different benefits\footnotemark\footnotetext{Written by Jose}
\section{Linear Filters}
With \textbf{Linear Filtering}, or specifically linear spatial filtering, the function which is used to pass the image through
must be linear and shift invariant. The function $\pmb{L}$ is considered linear if there exists two constants $a$ and $b$ for any
two input pixels $f_1(m, n)$ and $f_2(m, n)$ such that

@@

The desired output pixel $g(i, j)$, where $i$ and $j$ specify the
coordinates of it, is based on a $M \times N$ sized neighborhood, 
meaning not only does one pixel define an output pixel, but also a specified number
of its neighbors. The influence of each pixel in the neighborhood is defined by the filter coefficient $h(k, l)$,
also called its \emph{kernel} or \emph{mask}\cite{Szeliski_2022}. With this formula, it is possible to have a kernel that is rectangular and not square,
but through convention, it is generally preferred to have the kernel be square. As later discussed, square matrices are also more cost effective to calculate.

@@

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.30\textwidth]{codes/color.jpg}}
    \subfigure[shifted]{\includegraphics[width=0.30\textwidth]{img/colorShift.jpg}}
    \caption{Hue shifting using linear filters of the GALEX telescope image\cite{NASA_on_The_Commons_23}}
    \label{fig:hueshift}
\end{figure}\end{minipage}

The simplest application for linear filters are for smoothing(blurring) images. This not only serves their inherent purpose, but as later discussed, it can be used to aid other applications such as
edge detection or image enhancement.
\begin{itemize}
    \item \textbf{Box Filter:} each pixel is determined by the average of a $K \times K$ 
        neighborhood of pixels. It is also often called the \emph{moving average}, as the calculation can be done with scanlines, subtracting the
        old pixel and adding the newest pixels to the running sum. As the pixel is influenced equally by a square neighborhood, this may not be ideal as some `square artifacts' can be seen.
    \item \textbf{Bartlett Filter:} one solution to this is by separably convolving the image
        with a linear `tent' function (having its peak in the middle). A $3\times 3$ version of this is called the bilinear kernel.
    \item \textbf{Approximate Gaussian Filter:} if the Bartlett kernel is convolved with itself, this will produce an approximate version of the gaussian kernel.
\end{itemize}

This intuitively can be used to reduce the amount of noise on a picture. However, since it naively blurs the entire image,
details are also lost in the process.\\
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.32\textwidth]{codes/original_noise.jpg}}
    \subfigure[noise]{\includegraphics[width=0.32\textwidth]{codes/noise.jpg}}
    \subfigure[gaussian]{\includegraphics[width=0.32\textwidth]{img/noiseSuppressed.jpg}}
    \caption{(a) Original image of the Star City Museum\cite{NASA_on_The_Commons_14}(b) Noise added artificially (c) Blurred with Gaussian (k=27 $\sigma=7$)}
    \label{fig:noise}
\end{figure}\end{minipage}

@@

These filters are mainly used to smoothen(blur) the image, where the low frequencies are let through as well as reducing the effect of the high frequencies (high frequencies and low frequencies both referring to
the brightness/darkness of each pixel). In practical applications, this is done to reduce high frequency noises.
A creative application to smoothing the images, is that the smoothen images can be used as a mask to sharpen the image by adding the difference between the original and blurred image.
This method is called \emph{unsharp masking}, aiming to enhance the difference between the shadows and higlights on an image\cite{Szeliski_2022}.
\[
  g_{sharp} = f + \gamma(f - h_{blur} \ast f)  
\]
\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[original]{\includegraphics[width=0.19\textwidth]{img/original_mask.jpg}}
    \subfigure[unsharp mask]{\includegraphics[width=0.19\textwidth]{img/unsharp_mask.jpg}}
    \subfigure[difference]{\includegraphics[width=0.19\textwidth]{img/difference_unsharp.jpg}}
    \subfigure[sharpened]{\includegraphics[width=0.19\textwidth]{img/sharpened.jpg}}
    \caption{Unsharp masking with Gaussian blur}\label{fig:unsharp_mask}
\end{figure}\end{minipage}
The result of the image traces a bright outline througout the image between the darker(shadow) and lighter part. It can also be seen, that
the difference between the blurred image and the original results in an approximate edge detection. A better way to extract edges by means of linear filtering is
by the use of the \emph{Sobel filter}. \cite{Szeliski_2022}

@@

A similar to operator to the sobel filter is called the \emph{Prewitt operator}, which instead of convolving it with a perpendicular tent filter,
it instead uses an average operator. Commonly however, the Sobel operator is preferred as it provides better performance\cite{NIXON2020141}. Through testing, the Sobel Operator seems
to be less selective than the Prewitt Operator, but one can argue that the Prewitt is then preferred as it is more accurate.

\begin{minipage}{\textwidth}\begin{figure}[H]
    \centering
    \subfigure[Pattern]{\includegraphics*[width=0.29\textwidth]{codes/pattern.png}}
    \subfigure[Sobel]{\includegraphics[width=0.29\textwidth]{img/directionSobel.png}}
    \subfigure[Prewitt]{\includegraphics[width=0.29\textwidth]{img/directionScharr.png}}
    \caption{Vertical edge detection using Sobel and Prewitt on a test pattern}\label{fig:pattern}
\end{figure}\end{minipage}

As seen on Figure \ref{fig:pattern}, the Sobel filter tapers the edges at the top into a gradient, whereas the Prewitt marks it confidently without tapering. In practice however,
the minimal difference might not be worth the difference in the additional compute time \cite{NIXON2020141}.
